---
title: "John Hopkins COVID-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(tidyverse)
library(lubridate)
```

## Importing and Tidying COVID-19 Data

We will pull in and tidy the global data provided by the John Hopkins Whiting School of Engineering.

```{r COVID_data}
# Load our data from the source, https://github.com/CSSEGISandData/COVID-19/
base_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
filenames <- c("time_series_covid19_confirmed_US.csv",
                "time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_US.csv",
                "time_series_covid19_deaths_global.csv",
                "time_series_covid19_recovered_global.csv")
urls <- str_c(base_url, filenames)
# usa_cases <- read_csv(urls[0])
# usa_deaths <- read_csv(urls[1])
global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
global_recovered <- read_csv(urls[5])
```

```{r tidy_data, include=FALSE}
global_cases_clean <- global_cases %>%
    pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long),
                 names_to = "Date",
                 values_to = "Cases") %>%
    select(-c(Lat, Long)) %>%
    mutate(Date = mdy(Date))
sample_n(global_cases_clean, 5)

# `Province_State`, `Country_Region`, Lat, Long_, UID, code3, FIPS, iso2, iso3, Admin2, Combined_Key
global_deaths_clean <- global_deaths %>%
    pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long),
                 names_to = "Date",
                 values_to = "Deaths") %>%
    select(-c(Lat, Long)) %>%
    mutate(Date = mdy(Date))
sample_n(global_deaths_clean, 5)

global_recovered_clean <- global_recovered %>%
    pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long),
                 names_to = "Date",
                 values_to = "Recovered") %>%
    select(-c(Lat, Long)) %>%
    mutate(Date = mdy(Date))

sample_n(global_recovered_clean, 5)
```


## Visualizing Global COVID-19 Data



```{r total_data, echo = FALSE}

global_cases_daily <- global_cases_clean %>%
          group_by(Date) %>% 
          summarize(Cases = sum(Cases))

global_deaths_daily <- global_deaths_clean %>%
          group_by(Date) %>% 
          summarize(Deaths = sum(Deaths))

global_recovered_daily <- global_recovered_clean %>%
          group_by(Date) %>% 
          summarize(Recovered = sum(Recovered))

#sample_n(global_recovered_weekly, 5)

global_totals_by_country <- global_cases_clean %>%
                  full_join(global_deaths_clean) %>%
                  full_join(global_recovered_clean)

sample_n(global_totals_by_country, 5)

global_totals <- global_cases_daily %>%
                  full_join(global_deaths_daily) %>%
                  full_join(global_recovered_daily)

sample_n(global_totals, 5)

```

```{r graphing_global, echo = FALSE}

global_totals %>%
    filter(Cases > 0) %>%
    filter(Recovered > 0) %>%
    ggplot(aes(x = Date, y = Cases)) +
    geom_line(aes(y = Cases, color = "Cases")) +
    scale_y_log10() +
    geom_line(aes(y = Deaths, color = "Deaths")) +
    geom_line(aes(y = Recovered, color = "Recovered")) +
    theme(legend.title=element_blank(), legend.position="right", axis.text.x = element_text(angle = 45)) +
    labs(title = "COVID-19 Global Cases", y = NULL)

```
```{r population_data, echo = FALSE}

# Break down by the 10 most populous countries (https://www.census.gov/popclock/world)
top_countries <- c("China", "India", "US", "Indonesia", "Pakistan", "Brazil", "Nigeria", "Bangladesh", "Russia", "Mexico")
most_populous <- data.frame("Country_Region" = top_countries, "Population" = c(1397897720, 1339330514, 332475723, 275122131, 238181034, 219463862, 213445417, 164098818, 142320790, 130207371))

by_most_populous <- global_totals_by_country %>%
                    rename("Country_Region" = `Country/Region`) %>%
                    left_join(most_populous, by = "Country_Region") %>%
                    mutate(cases_per_thou = 1000* Cases / Population,
                           deaths_per_thou = 1000* Deaths / Population,
                           recov_per_thou = 1000* Recovered / Population)
                  
```


```{r graphing_countries, echo = FALSE}

by_most_populous %>%
    filter(cases_per_thou > 0.01) %>%
    ggplot(aes(x = Date, y = cases_per_thou)) +
    geom_line(aes(y = cases_per_thou, color = Country_Region)) + 
    scale_y_log10() +
    theme(legend.position="right", axis.text.x = element_text(angle = 90)) +
    labs(title = "COVID-19 Cases Per Thousand, by Country (10 Most Populous)", y = NULL) + 
    guides(color=guide_legend(title="Country/Region"))

```

## Modeling Recoveries per Cases, Over Time

Modeling over time recoveries per confirmed cases should represent improvements in recovering the infected. 

```{r modeling_recoveries, echo = FALSE}

recovery_model <- lm(Recovered ~ Deaths, data = global_totals)

# summary(recovery_model)
global_totals_with_pred <- global_totals %>% mutate(Prediction = predict(recovery_model))

global_totals_with_pred %>%
    filter(Cases > 0) %>%
    filter(Recovered > 0) %>%
    ggplot(aes(x = Date, y = Cases)) +
    geom_point(aes(y = Cases, color = "Cases")) +
    geom_point(aes(y = Recovered, color = "Recovered")) +
    geom_line(aes(y = Prediction, color = "Recovery Model")) +
    theme(legend.title=element_blank(), legend.position="right", axis.text.x = element_text(angle = 45)) +
    labs(title = "COVID-19 Global Cases \n(with Recovery Modeling)", y = NULL)

```

## Possible Bias

Population values were provided by the U.S. Census Bureau, a United States government data entity and population estimates may be skewed.

The John Hopkins CSSE COVID-19 data is aggregated from many data sources (https://github.com/CSSEGISandData/COVID-19), which might be misinterpreted as making it less bias. Each individual source could be potentially biased, particularly given that it is usually a government entity from these various countries reporting their own numbers. One example that has raised suspicion before are the case numbers China has reported - refer to the "Cases by Country" graph and note that China's infection rate was similar to other countries at first but quickly leveled out, a pattern not repeated any where else.


## Conclusion

Exploring the John Hopkins COVID-19 case data is of course alarming, cases and deaths seem to increase exponentially. However, seeing cases per thousand on a per country basis level out is encouraging. Additionally, modeling the global ratio of recoveries versus cases shows increasing improvements in medical treatment.


